---
title: "National Water Use example"
author: "Scott Worland"
date: "Thursday, July 16, 2015"
output:
  html_document:
    theme: cosmo
    toc: yes
---

### Preface

This code calculates the change in water use as a percentage of the average Mississippi River discharge. This is an example of how a commented script file will look.If a something requires a source, it is linked within the text.

### Input data

The first step is to load the [data](http://waterdata.usgs.gov/nwis/water_use), which is done with a simple R command,
```{r, echo=F, error=F,warning=F}
setwd("C:\\Users\\scworlan\\Documents\\Thermoelectric\\R_code\\Thermoelectric\\Water_use_example")
```

```{r, error=F, warning=F}
d = read.delim("USGS_national_WU_data.txt", header=T)
```

The next step is to manually include the [Mississippi River discharge](http://pubs.usgs.gov/of/1987/ofr87-242/),

```{r, eval=FALSE, warning=F}
Q1 = 593000 #ft3/s
```

These are the only input for this simple script, so I will next move into the calculations.

### Calculations

1. Convert the discharge data from ft$^3$/S into bgal/day. There are 7.48 gallons in 1 ft$^3$, so the calculation is straightforward,

$$
Q2 = (Q1 * 7.48 * 60 * 60 * 24)/1e9
$$

and is implemented in R below:

```{r, eval=FALSE, warning=F}
gal = 7.48 #gallons in ft3
Q2 = (Q1*gal*60*60*24)/1e9 #bgal/day
```

2. The next calculation is to find the annual differences for each water type. This is simply a column difference with lag = 1. This can be more cleanly implemented in R using an extra R package, but I just use base operations below:

```{r, eval=FALSE, warning=F}
d2 = as.matrix(t(d[,2:7])) #convert data.frame to matrix
d3 = t(diff(d2,lag=1)) # take difference of matrix
```

3. The final calculation is to calculate the annual changes as a percentage of the average discharge of the MS River. This is done by dividing each cell of the difference matrix, d3, by the discharge, Q2:

$$
d4 = (d3/Q2) * 100
$$

```{r, eval=FALSE, warning=F}
d4 = round(d3/Q2*100, 1)
```

### Export results

The final step is to clean up the data and export to excel:
```{r, eval=FALSE, warning=F}
library(xlsx)
out = data.frame(d[,1],d4)
colnames(out) = c("Type","1985-1990","1990-1995","1995-2000","2000-2005","2005-2010")
write.xlsx(out,file = "WU_output_R.xlsx",row.names = FALSE)
```




~ ~ ~ ~ ~


